---
title: "Vis 1"
author: "Chhiring Lama"
date: "2024-09-26"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(rnoaa)
library(ggridges)
```

```{r}
weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USW00022534", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2021-01-01",
    date_max = "2022-12-31") |>
  mutate(
    name = recode(
      id, 
      USW00094728 = "CentralPark_NY", 
      USW00022534 = "Molokai_HI",
      USS0023B17S = "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10,
    month = lubridate::floor_date(date, unit = "month")) |>
  select(name, id, everything())

```


Making our first plot
```{r}
ggplot(weather_df, aes(x = tmin, y = tmax)) +
  geom_point()
```

We can save and print it later
```{r}
gg_weather_scatterplot <- ggplot(weather_df, aes(x = tmin, y = tmax)) +
  geom_point()

gg_weather_scatterplot
```

Check why some rows are missing
```{r}

```


## fancier scatterplots

```{r}
weather_df |> 
  ggplot(aes(x = tmin, y = tmax, color = name)) +
  geom_point(alpha = 0.6, size = 0.8) +
  geom_smooth(se = FALSE)
```

Where we define the `aes` matters. The smoothing looks different when used in `geom` where we only define the color grouping only for the scatterplot, the grouping is not global so `geom_smooth` does not recognize this. 
```{r}
weather_df |> 
  ggplot(aes(x = date, y = tmax)) +
  geom_point(alpha = 0.6, size = 0.8, aes(color = name)) +
  geom_smooth(se = FALSE)
```

Use facet
```{r}
weather_df |> 
  ggplot(aes(x = tmin, y = tmax)) +
  geom_point(alpha = 0.6, size = 0.8, aes(color = name)) +
  geom_smooth(se = FALSE) +
  facet_grid(.~name)
```

```{r}
weather_df |> 
  ggplot(aes(x = date, y = tmax, color = name, size = prcp)) +
  geom_point(alpha = 0.6) +
  geom_smooth(se = FALSE) +
  facet_grid(.~name)
```

## Learning Assessment
Write a code chain that starts with weather_df; focuses only on Central Park, converts temperatures to Fahrenheit, makes a scatterplot of min vs. max temperature, and overlays a linear regression line (using options in geom_smooth())

### Solution
```{r}
weather_df |> 
  filter(name == "CentralPark_NY") |> 
  mutate(tmax_fahrenheit = ((tmax * 9/5) + 32), 
         tmin_fahrenheit = ((tmin * 9/5) + 32)) |> 
  ggplot(aes(x = tmin_fahrenheit, y = tmax_fahrenheit, size = prcp)) +
  geom_point(alpha = 0.6) +
  geom_smooth(se = FALSE, method = "lm")

```

We can look at the density at certain point in the graph too.
```{r}
weather_df |> 
  ggplot(aes(x = tmin, y = tmax)) +
  geom_hex()
```

## Learning Assessment
In the preceding, we set the alpha aesthetic “by hand” instead of mapping it to a variable. This is possible for other aesthetics too. To check your understanding of this point, try to explain why the two lines below don’t produce the same result:

```{r}
weather_df |> 
  ggplot(aes(x = tmin, y = tmax)) +
  geom_point(color = "blue")

weather_df |> 
  ggplot(aes(x = tmin, y = tmax,color = "blue")) +
  geom_point()
```

### Solution:
In the second case, because we put the color in aesthetic which take variables, it created a variable whose value is "blue" for all cases, and ends up coloring with a single color. The primary chose for R is red. 

## Univariate plots

```{r}
weather_df |> 
  ggplot(aes(x = tmin, fill = name)) +
  geom_histogram(position = "dodge", binwidth = 2)
```

How we can fix this:
```{r}
weather_df |> 
  ggplot(aes(x = tmin, fill = name)) +
  geom_histogram(position = "dodge", binwidth = 2) +
  facet_grid(.~name)
```

We can use density plot instead
```{r}
weather_df |> 
  ggplot(aes(x = tmin, fill = name)) +
  geom_density(alpha = .3)
```

Using box plot
```{r}
weather_df |> 
  ggplot(aes(x = name, y = tmin, fill = name)) +
  geom_boxplot(alpha = .3)
```

A variation of box plot: violin plot (this helps identify difference between multiple groups, and also identify bimodal data distribution)
```{r}
weather_df |> 
  ggplot(aes(x = name, y = tmin, fill = name)) +
  geom_violin(alpha = 0.3) + 
  stat_summary(fun = "median", color = "blue")
```

Another variation: ridge plot
```{r}
weather_df |> 
  ggplot(aes(x = tmax, y = name)) + 
  geom_density_ridges(scale = .85)
```

## Learning Assessment: 
Make plots that compare precipitation across locations. Try a histogram, a density plot, a boxplot, a violin plot, and a ridgeplot; use aesthetic mappings to make your figure readable

### Solution
```{r}
ggplot(weather_df, aes(y = prcp, fill = name)) +
  geom_histogram(position = "dodge", alpha = 0.8, binwidth = 50) +
  facet_grid(name~.)
```

```{r}
ggplot(weather_df, aes(y = prcp, fill = name)) +
  geom_density(alpha = 0.8)
```

can log-transform the density plot
```{r}
ggplot(weather_df, aes(y = prcp, fill = name)) +
  geom_density(alpha = 0.8) +
  scale_y_log10()
```

```{r}
ggplot(weather_df, aes(x = name, y = prcp, fill = name)) +
  geom_boxplot(alpha = 0.8) 
```

```{r}
ggplot(weather_df, aes(x = name, y = prcp, fill = name)) +
  geom_violin(alpha = 0.8) + 
  stat_summary(fun = "median", color = "blue")
```

```{r}
ggplot(weather_df, aes(y = name, x = prcp, fill = name)) +
  geom_density_ridges(scale = .85) 
```

So far, the box plot lets us look at the distribution and identify the outliers. 

## Saving and embedding plots
```{r}
gg_weather = 
  weather_df |> 
  ggplot(aes(x = date, y = tmax, color = name)) +
  geom_point()

ggsave("plots/ggp_weather.pdf", gg_weather, height = 5, width = 6)
```

embedding plots
```{r, fig.width = 8, fig.asp = .6, out.width = "90%"}
weather_df |> 
  ggplot(aes(x = date, y = tmax, color = name)) +
  geom_point()
```

